# Enter the Blank Person! — Coding Multi-Level Categorical Predictors {#chap-4}

```{r}
#| echo: false
load("data/chapter3_env.RData")  # Inherit all objects from Chapter 3
```

Back when I was in elementary school, a wildly popular game swept through my peers. Its Chinese title carried an oddly charming cadence, *Children Fighting Together 2* (《小朋友齐打交2》), known internationally as *Little Fighter 2* (<https://www.lf2.net/>). I recently searched online and was delighted to find that an official-looking website still lingers on the web. I warmly recommend curious readers try it out (though do remember to come back and finish this book).

I bring up this arcade classic for two reasons: first, it was ridiculously fun; second, among its colorful roster of martial artists, there was a peculiar character whom I always called the "Blank Person" (Template). Unlike the other warriors, the Blank Person possessed no flashy magical abilities or special combo moves—only bare-knuckle punches and kicks. Amidst the vibrant cast, this minimalist, monochrome figure felt delightfully stark and out of place.

![](images/clipboard-2108693482.bmp)

Cruel as it sounds, in research design, such a Blank Person is the born-to-be, consummate control group. Think about it: we already have Helicopter People and Bag People, and we wish to evaluate their magical talent. What we urgently need is a cohort of people who possess zero gender self-identity, yet are functional human beings with all standard baseline capabilities. That way, we can isolate and evaluate the unique impact of each identity on magical talent.

Without further ado, let us add a cohort of Blank People into our ongoing dataset:

```{r}
## Generate data for 50 Blank People
set.seed(234)
df.blank <- data.frame(ID = c(101:150), iden = c(rep("Blank Person", 50)))
df.blank$magic_score <- rnorm(n = 50, mean = 95, sd = 10)

## Combine Blank People with our existing Helicopter and Bag People data
df4.1 <- df4 <- rbind(df1, df.blank)

## Verify the combined data
tail(df4)
summary(df4)
```

Let's inspect their descriptive distributions:

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-responsive: true
#| out-width: "100%"

library(ggplot2)

gg <- ggplot2::ggplot(df4.1, aes(x = iden, y = magic_score, fill = iden)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.2, color = "black",
               position = position_dodge(0.9),
               outlier.shape = NA,
               show.legend = FALSE) +
  geom_jitter(width = 0.1, alpha = 0.3, color = "black", show.legend = FALSE) +
  stat_summary(
    fun = mean, geom = "point",
    shape = 17, size = 5, color = "red",
    position = position_dodge(0.9),
    show.legend = FALSE
  ) +
  scale_color_manual(values = c("#69b3a2", "#ff9f40", "purple")) +
  guides(color = "none") +
  scale_fill_manual(values = c("#69b3a2", "#ff9f40", "purple")) +
  labs(title = "Magic Talent Scores across Three Tribes", x = "Identity", y = "Magic Score") + 
  theme_minimal() +
  theme(aspect.ratio = 3/1,
        plot.title = element_text(hjust = 0.5))

plotly::ggplotly(gg)
```

Now our data is assembled, and it is time for analysis. What is our research hypothesis? Following our logic from earlier chapters, a natural hypothesis is that differences in magical talent exist among Helicopter People, Bag People, and Blank People:

$$\mu_{\text{Helicopter}} \ne \mu_{\text{Bag}} \ne \mu_{\text{Blank}}$$

The corresponding null hypothesis states that all three groups share identical magical talent:

$$H_0: \mu_{\text{Helicopter}} = \mu_{\text{Bag}} = \mu_{\text{Blank}}$$

## One-Way ANOVA with Three Levels in Action {#sec-onewayanova}

Any seasoned student will recognize this immediately as a classic one-way Analysis of Variance (ANOVA). While we used ANOVA in Chapter 3 to compare two groups, real-world textbooks typically showcase factors with three or more levels—precisely because ANOVA shines where the $t$-test cannot tread, evaluating omnibus differences across three or more groups simultaneously.

Let's run the ANOVA:

```{r}
## Disregard my variable naming habits; naming is governed purely by mood
my.aov4.1 <- aov(magic_score ~ iden, df4.1)
## We prefer reporting via car::Anova; reasons will be discussed later
car::Anova(my.aov4.1)
```

The omnibus test is highly significant: $F(2, 147) = 35.415, p = 2.794 \times 10^{-13}$.

However, the primary mission of this book is not merely to march you through standard ANOVA steps. In the previous chapter, we established that ANOVA and linear regression are mathematical siblings under the General Linear Model. Can we test this three-level hypothesis within a linear regression framework?

## Coding: The Magic of Transforming Categories into Continuous Predictors

To analyze differences among three groups via linear regression, our predictor is an unordered categorical variable with three levels. The primary hurdle is: how do we feed a multi-level categorical variable into a linear regression model? We touched upon this in Chapter 3 (@tip-coding1 and @tip-coding2), but back then our factor had only two levels. Now, with the intrusion of the Blank Person, our factor has three levels. What should we do?

In Chapter 2 and 3, we assigned numeric values to our two groups: Helicopter = 0, Bag = 1. A naive, instinctive impulse is simply to extend that sequence: let Blank Person = 2. Will this work? Let's test it:

```{r}
df4.1$iden_n <- c(rep(0, 50), rep(1, 50), rep(2, 50))
my.lm4.1 <- lm(magic_score ~ iden_n, df4.1)
summary(my.lm4.1)
```

Evidently, it does not. In Chapter 3, we proved that model fit in regression should match the ANOVA omnibus test. Yet here, the regression model reports $F(1, 148) = 43.41, p = 7.296 \times 10^{-10}$—completely at odds with our ANOVA result from @sec-onewayanova ($F = 35.415$). Where did things go wrong?

As emphasized earlier, a two-level categorical variable can be treated as a special "interval" variable because two points define exactly one distance. But with three groups, there are three pairwise distances: Helicopter-to-Bag, Bag-to-Blank, and Helicopter-to-Blank. Are these three distances equal? Absolutely not! Assigning $\{0, 1, 2\}$ artificially forces the distance between Helicopter and Blank to be exactly twice the distance between Helicopter and Bag. This unwarranted geometric constraint distorts the model.

In @tip-coding2, we noted that regardless of how a binary variable is coded, OLS regression always predicts the respective group means. What about this model?

```{r}
table(predict(my.lm4.1))
```

Compare these predictions with our empirical sample means:

```{r}
group_means <- aggregate(magic_score ~ iden, data = df4.1, FUN = mean)
group_means
```

They are not even remotely close! Treating three categorical levels as sequential integers prevents the regression model from recovering the actual group means.

We clearly require a principled coding scheme—one that grants multi-level factors the mathematical flexibility to enter linear regression as proper numeric predictors.

### 1. Dummy Coding

To make regression work properly, model predictions must correspond to empirical group means. Let's revisit our two-level model from Chapter 3:

```{r}
summary(lm1)
```

Our regression equation was $\hat{y} = 81.240 + 14.055x$. At $x = 0$, $\hat{y} = b_0 = 81.240$ (the Helicopter mean). At $x = 1$, $\hat{y} = b_0 + b_1 = 95.295$ (the Bag mean).

What is missing now? The mean of the Blank People ($94.37$). How can we incorporate it? Sequential integers failed because categories are not equispaced. But categorical variables *are* effectively equispaced when restricted to two levels. So what if we introduce an additional indicator variable?

Suppose we keep $D_1$ (Helicopter = 0, Bag = 1, capturing the difference between Bag and Helicopter via $b_1$), and create a second variable $D_2$ (Helicopter = 0, Blank = 1, capturing the difference between Blank and Helicopter via $b_2$):

| Identity | $D_1$ (Bag vs. Heli) | $D_2$ (Blank vs. Heli) |
|---|:---:|:---:|
| Helicopter | 0 | 0 |
| Bag | 1 | — |
| Blank | — | 1 |

If we ran separate bivariate regressions for $D_1$ and $D_2$, that would merely replicate two isolated $t$-tests. We want a single omnibus model containing both predictors simultaneously:

$$\hat{y} = b_0 + b_1 D_1 + b_2 D_2$$

How should we code the Blank People on $D_1$, and Bag People on $D_2$?
For Helicopter People ($D_1 = 0, D_2 = 0$):
$$\hat{y}_{\text{Heli}} = b_0$$ {#eq-dc0}

For Bag People ($D_1 = 1, D_2 = x$):
$$\hat{y}_{\text{Bag}} = b_0 + b_1 + b_2 x$$
To isolate $b_1$, we simply set $x = 0$:
$$\hat{y}_{\text{Bag}} = b_0 + b_1$$ {#eq-dc1}

Likewise for Blank People on $D_1$, setting its code to 0 yields:
$$\hat{y}_{\text{Blank}} = b_0 + b_2$$ {#eq-dc2}

Behold! Our coding matrix is complete:

| Identity | $D_1$ (Bag vs. Heli) | $D_2$ (Blank vs. Heli) |
|---|:---:|:---:|
| Helicopter | 0 | 0 |
| Bag | 1 | 0 |
| Blank | 0 | 1 |

Let's test this in R:

```{r}
## Construct D1 and D2 according to the coding matrix
df4.1$D1 <- ifelse(df4.1$iden == "Plastic Bag", 1, 0)
df4.1$D2 <- ifelse(df4.1$iden == "Blank Person", 1, 0)

## Fit the regression model
my.lm4.2 <- lm(magic_score ~ D1 + D2, df4.1)
summary(my.lm4.2)
```

Compare this with our ANOVA output:

```{r}
car::Anova(my.aov4.1)
```

What a coincidence! The overall model $F$-test matches the ANOVA omnibus test down to the last decimal: $F(2, 147) = 35.415, p = 2.794 \times 10^{-13}$!

Now examine the regression coefficients against our empirical group means:

```{r}
group_means
```

The intercept $b_0 = 81.240$ is the mean of Helicopter People. From @eq-dc0, @eq-dc1, and @eq-dc2:

$$\begin{align}
b_0 &= \hat{y}_{\text{Heli}} = 81.240 \\
b_1 &= \hat{y}_{\text{Bag}} - \hat{y}_{\text{Heli}} = 95.294 - 81.240 = 14.055 \\
b_2 &= \hat{y}_{\text{Blank}} - \hat{y}_{\text{Heli}} = 94.370 - 81.240 = 13.131
\end{align}$$

Testing $b_1 = 0$ evaluates whether Bag People differ from Helicopter People; testing $b_2 = 0$ evaluates whether Blank People differ from Helicopter People.

Notice that although $b_1 = 14.055$ matches the mean difference from Chapter 2, its significance test ($t = 7.523, p = 4.91 \times 10^{-12}$) differs slightly from Chapter 2's isolated $t$-test:

```{r}
t.test(magic_score ~ iden, df1.1, var.equal = TRUE)
```

Why? Because the regression model calculates standard error by pooling residuals across *all three groups* (including Blank People). This is the hallmark of a **planned contrast**.

::: {#tip-planned-posthoc .callout-note icon="false"}
## Extension: Are Planned Comparisons Follow-Ups to One-Way ANOVA? What about Post-Hoc Comparisons?

Many statistics courses follow a familiar sequence: first teach or conduct a one-way ANOVA, then teach or conduct comparisons among its levels. As a result, I used to think—and I know other people do too—that ANOVA is the starting point of all analysis: after the omnibus test, we begin to further examine its results. For example, having found a significant difference in magical talent among Helicopter People, Bag People, and Blank People, we might then ask whether Helicopter People differ from Bag People, whether Bag People differ from Blank People, and whether Helicopter People differ from Blank People.

Are these not exactly the hypotheses we can test using dummy coding? Moreover, the set of dummy-coded predictors can jointly test whether the three groups differ at all. Does this mean these planned comparisons, taken together, *are* one-way ANOVA—and that after a significant ANOVA we can use them to explain its results?

A statistically savvy reader will surely say: no, that is not right. Once the ANOVA is significant, what we should conduct is a post-hoc comparison, not a planned comparison.

So, is a post-hoc test a further analysis of the one-way ANOVA result?

In terms of workflow, yes. Looking at the words *planned* and *post-hoc*,[^chap4-1] a post-hoc test is conducted after the ANOVA—that is what "post-hoc" means. This also meets a reasonable empirical need: after discovering a significant omnibus difference among the levels of a factor, we naturally want to know exactly which levels differ; simply knowing that some difference exists is usually not enough. That further analysis is a post-hoc test.


A statistically savvy student may object again: isn't the defining feature of a post-hoc test that it controls the Type I error rate?

That is true of many post-hoc procedures. But logically, a post-hoc test does not *have* to control Type I error. Consider Fisher's LSD (*Least Significant Difference*), a post-hoc test that is a variation on the $t$-test. What is varied? We know that the numerator of a $t$-statistic is a mean difference and its denominator is the standard error of the sampling distribution. The mean difference itself is straightforward—for example, the mean for Helicopter People minus the mean for Bag People. In a one-way design with several groups, however, the standard error can take the other groups (such as Blank People) into account. The LSD standard error is:

$$SE = \sqrt{MS_{\text{Within}}\left(\frac{1}{n_i} + \frac{1}{n_j}\right)}$$

The independent-samples $t$-test can be regarded as a special case; with three or more groups, the procedure uses them to estimate a common standard error. What does that look like in our example?

```{r}
DescTools::PostHocTest(my.aov4.1, method = "lsd")
```

Compare these results with the dummy-coding results above. Notice anything? The $p$-values are the same. Once again, we have inadvertently touched the elephant and glimpsed the common ground among LSD, the $t$-test, linear regression, and ANOVA. This also answers the question we raised in this section: compared with the isolated independent-samples $t$-test, our planned comparison takes the standard error contributed by the other groups into account.

Wait, weren't we discussing post-hoc comparisons? How did we end up back at planned comparisons? Congratulations—you spotted the key point. Fisher's LSD and the set of all possible two-group planned comparisons are essentially the same tests. The difference is simply whether they are conducted after the ANOVA.

It is not entirely wrong to say that post-hoc tests are characterized by Type I error control. Fisher's LSD is largely retained in textbooks but has been abandoned in much empirical practice precisely because it does not control Type I error. Why does that control matter?[^chap4-2]


Suppose we have found a difference among the three groups and then conduct three separate $t$-tests using $\alpha = 0.05$. If each test is treated separately, the probability that none makes a Type I error is $0.95^3$. Therefore, the probability of at least one Type I error across the study is:

```{r}
1 - 0.95^3  # The probability that all three tests avoid a Type I error is 0.95^3
```

Our expected level is 0.05, but this has nearly tripled it. What happens if there are more levels? The number of comparisons appears in the exponent, so the problem can become severe. Repeated hypothesis tests without Type I error control greatly increase the chance of a false positive. Thus, in the common situation where we need to perform many tests—that is, in post-hoc testing—controlling Type I error is both necessary and important.[^chap4-3]


The many different post-hoc procedures exist because we need to balance Type I error inflation against the loss of statistical power caused by stricter control. I do not want to spend too much time on post-hoc methods: there is no universally correct answer, and this is not the main purpose of the book. Let us look at one of the strictest and most intuitive methods: Bonferroni. Its logic is straightforward. If conducting $c$ tests inflates the family-wise Type I error rate, limit the per-test alpha by dividing 0.05 equally among them: $\alpha_{\text{per test}} = 0.05/c$.

We can calculate the family-wise rate in the same way. It stays around 0.05—quite strict, indeed. But who would run 999 tests? Bonferroni is often the post-hoc procedure that most effectively heads off objections.

```{r}
1 - (1 - 0.05 / 3)^3
1 - (1 - 0.05 / 9)^9
1 - (1 - 0.05 / 99)^99
1 - (1 - 0.05 / 999)^999
```

Now let's see the procedure in practice:

```{r}
DescTools::PostHocTest(my.aov4.1, method = "bonferroni")
```

The reported results have adjusted $p$-values, but the underlying tests are equivalent. We discussed a related issue when considering one-tailed versus two-tailed $p$-values in Chapter 2. In our example, there are three comparisons, so multiplying the LSD $p$-values by 3 gives the Bonferroni-adjusted results (with adjusted $p$-values capped at 1):

```{r}
DescTools::PostHocTest(my.aov4.1, method = "lsd")$iden[, "pval"] * 3
DescTools::PostHocTest(my.aov4.1, method = "bonferroni")$iden[, "pval"]
```

Now, back to planned comparisons. Does a planned test mean a comparison conducted before ANOVA?

No. We need to ask a different question: is a difference between Helicopter People and Bag People part of the broader question of whether the three groups differ? Are these questions necessarily sequential? Must we first know whether all three groups differ before we can ask whether Helicopter People and Bag People differ? If the only populations in the world were Helicopter People, Bag People, and Blank People, and I had data from all three but cared only about the difference between the first two, could I ask that question?

Mathematically, differences among all three groups and a difference between two particular groups are related. But remember that research uses statistics to test hypotheses. Planned comparisons and omnibus ANOVA may be mathematically entangled; in our research, however, we should focus on the hypothesis we have and how to test it. A planned comparison is not a subordinate part of ANOVA. It can be conducted before or after ANOVA, and it is entirely legitimate to conduct a planned comparison without running ANOVA at all. The key is whether the planned test evaluates a hypothesis motivated by the theory and research question—not whether we remembered to use it only after seeing a significant ANOVA.
:::

In dummy coding, for a factor with $k$ levels, we construct $k - 1$ indicator variables.[^chap4-3] We select one level as the baseline (coded as 0 across all dummy variables), and assign a 1 to each remaining level on its designated indicator:


| Factor Level | $D_1$ | $D_2$ | $\dots$ | $D_{k-1}$ |
|:---:|:---:|:---:|:---:|:---:|
| Reference Baseline | 0 | 0 | $\dots$ | 0 |
| Level 2 | 1 | 0 | $\dots$ | 0 |
| Level 3 | 0 | 1 | $\dots$ | 0 |
| $\dots$ | $\dots$ | $\dots$ | $\dots$ | $\dots$ |
| Level $k$ | 0 | 0 | $\dots$ | 1 |

At the start of this chapter, we introduced Blank People as a natural control group. Let's place Blank People at the intercept baseline ($D_1 = D_2 = 0$):

| Identity | $D_{1n}$ (Bag vs. Blank) | $D_{2n}$ (Heli vs. Blank) |
|---|:---:|:---:|
| Helicopter | 0 | 1 |
| Bag | 1 | 0 |
| Blank (Baseline) | 0 | 0 |

```{r}
df4.1$D1n <- ifelse(df4.1$iden == "Plastic Bag", 1, 0)
df4.1$D2n <- ifelse(df4.1$iden == "Attack Helicopter", 1, 0)

my.lm4.3 <- lm(magic_score ~ D1n + D2n, df4.1)
summary(my.lm4.3)
```

The omnibus model fit remains identical ($F = 35.41$). Intercept $b_0 = 94.37$ is the mean of Blank People. The slope $D_{1n} = 0.924, p = 0.622$ shows that Bag People do not significantly differ from Blank People, whereas Helicopter People score significantly lower ($D_{2n} = -13.131, p = 7.26 \times 10^{-11}$).[^chap4-4] Identifying as a Helicopter suppresses magical talent relative to baseline, whereas identifying as a Plastic Bag offers no special magical boost!

::: {#tip-dcal .callout-tip icon="false"}
## Food for Thought: What Happens If We Do Not Code the Blank People in $D_1$ or the Bag People in $D_2$ as 0?

As the title asks: what happens if we do not follow the dummy-coding rule? Let's make a small change and use the following coding matrix instead:

| Identity | $D_1$ | $D_2$ |
|---|:---:|:---:|
| Helicopter | 0 | 0 |
| Bag | 1 | 0 |
| Blank | 1 | 1 |

I have generously written out the implementation for you:

```{r}
df4.1$R1 <- c(rep(0, 50), rep(1, 50), rep(1, 50))
df4.1$R2 <- c(rep(0, 50), rep(0, 50), rep(1, 50))
my.lm4.4 <- lm(magic_score ~ R1 + R2, df4.1)
summary(my.lm4.4)
```

The question is: what hypotheses do these two regression coefficients test? Thoughtful readers can work it out from the coding matrix and verify their answers algebraically.
:::

### 2. Effects Coding (Deviation Coding)

At the start of this chapter, we praised Blank People as an ideal control group. But in reality, control groups are rarely pristine. Suppose that one day, the Blank People achieve political consciousness and declare that their blank identity is not an "absence of identity," but an affirmative identity in its own right. Those treacherous Blank People have sabotaged our control group!

In real-world research—especially with demographic variables like ethnicity or nationality—a pure control group rarely exists. How can we test how each group performs relative to the *average* across all groups?[^chap4-6]

The Blank People may joke that their identity is "nothing," but nothing is still a kind of something. In the language of the old Buddhist verse, *"The Bodhi tree is not a tree; the bright mirror is not a stand"*: absence itself can be treated as a meaningful presence rather than a void. In less mystical terms, Blank is not necessarily the absence of an identity; it can be one identity among several. Our supposedly perfect control group has therefore escaped the control group!

This is more than a joke. When the variable of interest is a real-world demographic characteristic that researchers cannot manipulate, finding a clean control group with sufficient sample size may be impossible. In that situation, comparing each group with the overall average can be more meaningful than comparing each group with one arbitrarily selected baseline.

The word *average* suggests a natural question: why not compare each group's mean magical-talent score with the grand mean across all three groups? If you think about it, the between-group variation in ANOVA—the deviations of the group means from the grand mean—is already built around this idea.

With dummy coding and Helicopter People as the reference group, the regression coefficients describe the three means this way:

$$\begin{align}
b_0 &= \text{mean of Helicopter People} \\
b_1 &= b_0 + (\text{Bag mean} - \text{Helicopter mean}) = \text{Bag mean} \\
b_2 &= b_0 + (\text{Blank mean} - \text{Helicopter mean}) = \text{Blank mean}
\end{align}$$

We can change the reference point. Instead of using the Blank mean as a baseline, let the intercept be the grand mean, then express the Bag and Helicopter means as deviations from it:

$$\begin{align}
b_0 &= \text{grand mean} \\
b_1 &= \text{grand mean} + (\text{Bag mean} - \text{grand mean}) \\
b_2 &= \text{grand mean} + (\text{Helicopter mean} - \text{grand mean})
\end{align}$$

This way, testing the regression slopes tests whether each group's mean differs from the overall mean.

In effects coding, we set the baseline to the **unweighted grand mean** of all groups:

$$\begin{align}
b_0 &= (\hat{y}_{\text{Heli}} + \hat{y}_{\text{Bag}} + \hat{y}_{\text{Blank}}) / 3 \\
b_1 &= \hat{y}_{\text{Bag}} - b_0 \\
b_2 &= \hat{y}_{\text{Heli}} - b_0
\end{align}$$ {#eq-effect}

Solving for the group predictions in terms of coefficients:[^chap4-7]


$$\begin{align}
\hat{y}_{\text{Bag}} &= b_0 + b_1 \\
\hat{y}_{\text{Heli}} &= b_0 + b_2 \\
\hat{y}_{\text{Blank}} &= b_0 - b_1 - b_2
\end{align}$$

The algebra is not difficult, but it is worth doing by hand. We first substitute $b_1 = \hat{y}_{\text{Bag}} - b_0$ and $b_2 = \hat{y}_{\text{Heli}} - b_0$ into the equation for $b_0$. Solving the resulting system gives the three group predictions above. This is exactly the kind of derivation I recommend whenever you design a custom coding scheme: do not trust the labels alone—verify what each coefficient actually tests.

For clarity, the substitution step is:

$$b_0 = \frac{(b_0+b_2)+(b_0+b_1)+\hat{y}_{\text{Blank}}}{3}$$

Solving these equations yields the predictions above. In this coding matrix, the reference group receives $-1$ on both coding variables. This is the key distinction from dummy coding, where the reference group receives 0 on every coding variable.

To achieve this in regression, our coding matrix must assign $-1$ to the reference group:

| Identity | $E_1$ | $E_2$ |
|---|:---:|:---:|
| Helicopter | 0 | 1 |
| Bag | 1 | 0 |
| Blank (Reference) | -1 | -1 |

Let's test effects coding in R:

```{r}
df4.1$E1 <- ifelse(df4.1$iden == "Plastic Bag", 1, 
                   ifelse(df4.1$iden == "Blank Person", -1, 0))
df4.1$E2 <- ifelse(df4.1$iden == "Attack Helicopter", 1, 
                   ifelse(df4.1$iden == "Blank Person", -1, 0))

my.lm4.5 <- lm(magic_score ~ E1 + E2, df4.1)
summary(my.lm4.5)
```

Look at the coefficients:
- Intercept $b_0 = 90.3014$: the unweighted grand mean across all three tribes ($81.24 + 95.29 + 94.37) / 3$.
- $E_1 = 4.9930$: the deviation of Bag People from the grand mean ($95.294 - 90.301 = +4.993$).
- $E_2 = -9.0618$: the deviation of Helicopter People from the grand mean ($81.240 - 90.301 = -9.062$).

Both coefficients are statistically significant, proving that both Bag and Helicopter People diverge significantly from the general average.

The interpretation always depends on what the baseline represents. If the groups were three countries, for example, the "grand mean" here would mean the average of those three countries—not the average of the entire world. And if you want to see the Blank People's deviation from the grand mean, you can choose a different coding matrix and make Blank the group represented by one of the coefficients.

```{r}
mean(df4.1$magic_score)
group_means
```

The Blank People's deviation can be obtained by changing the reference coding matrix and choosing a different baseline. Effects coding does not make one group disappear; it changes the comparison target from a particular baseline group to the grand mean.

### 3. Contrast Coding

Perhaps some readers are wondering: are dummy coding and effects coding the only available options? Of course not, dear reader—there is more! We are about to introduce a wonderfully versatile, surprisingly slimming coding gadget: **contrast coding**.

When we introduced dummy coding, we noted that each coding variable can represent a planned comparison between two levels of a categorical factor. But readers familiar with planned comparisons may object: a planned comparison does not have to compare only two individual levels. For example, we might want to compare the combined group of Helicopter and Bag People with the Blank People. How can dummy coding represent that hypothesis? This is where contrast coding comes in.

Some readers may find the phrase "the difference between the two groups of Helicopter and Bag People and the single group of Blank People" a little confusing. To make it more concrete, imagine a parallel universe in which the identities of Helicopter and Bag People arise from a kind of animistic religion that developed into a secular culture. People form an identity by finding a concrete object in the world and developing a spiritual connection with it. Blank People, by contrast, have an identity that grew out of anime and other two-dimensional culture. They believe that the human mind originates in nothingness; they try to strip away irrelevant cultural imagery and seek the simplest possible basis for identity. For them, that basis is "blank."[^chap4-8] As researchers, we might therefore want to compare whether these two different cultural origins are associated with different magical talents.


Mathematically, how do we express such a hypothesis? If we compare two groups, our hypothesis might be:

$$\mu_{\text{Heli}} \ne \mu_{\text{Bag}}$$

By analogy, to compare Helicopter and Bag People together against Blank People, we can use the average of the first two population means and compare it with the third:[^chap4-9]


$$\frac{\mu_{\text{Heli}} + \mu_{\text{Bag}}}{2} - \mu_{\text{Blank}} \ne 0$$

Some researchers dislike having fractions in an expression, so we can multiply both sides by 2 without changing whether the contrast equals zero:

$$\mu_{\text{Heli}} + \mu_{\text{Bag}} - 2\mu_{\text{Blank}} \ne 0$$

More generally, a planned comparison among $k$ factor levels can be written as a weighted linear combination of population means:

$$L = w_1\mu_1 + w_2\mu_2 + \dots + w_k\mu_k \ne 0$$

For our first hypothesis, the weights are $w_1=1$, $w_2=1$, and $w_3=-2$. This notation may look familiar. In dummy coding and effects coding, do we not also multiply each group mean by a number and combine the results? Could we use those weights as the coding values? For the first contrast, the coding column would be:

| Identity | $C_1$ |
|---|:---:|
| Helicopter | 1 |
| Bag | 1 |
| Blank | -2 |

You could instead use the weights $w_1=\frac12$, $w_2=\frac12$, and $w_3=-1$. These weights express the same null hypothesis, because they differ only by multiplication by a non-zero constant. I encourage you to try both versions and compare their coefficients and standard errors.

How should we code the second contrast column? Suppose we also want to compare Helicopter People with Bag People, without involving Blank People:

$$\mu_{\text{Heli}} - \mu_{\text{Bag}} - 0\mu_{\text{Blank}} \ne 0$$

The weights are now $w_1=1$, $w_2=-1$, and $w_3=0$. Our two-column coding table becomes:

| Identity | $C_1$ | $C_2$ |
|---|:---:|:---:|
| Helicopter | 1 | 1 |
| Bag | 1 | -1 |
| Blank | -2 | 0 |

The columns encode two theoretically motivated comparisons. We can now enter them together into a regression model and test those hypotheses using the corresponding coefficient tests. The remaining issue is whether the columns interfere with one another; that is the question of orthogonality discussed below.


::: {.callout-tip}
## Food for Thought: A Half-Formed Discussion of Planned and Post-Hoc Comparisons

In @tip-planned-posthoc, we discussed how planned and post-hoc comparisons differ and when each may be appropriate. Having introduced contrast coding, we now have a much richer toolkit for planned comparisons: in principle, we can test any linear comparison of the means of a multi-level factor.

But that is not the question I want to discuss here. The question came from a post on a social-media platform.[^chap4-10] In broad outline, someone described a study with a five-level independent variable: one control group, which we will call $A$, and four treatment groups, $B$, $C$, $D$, and $E$. The researcher is interested only in whether each treatment group differs from the control group. Can they simply run four independent-samples $t$-tests: $A-B$, $A-C$, $A-D$, and $A-E$?

Dear thoughtful reader, pause here and decide for yourself: is that acceptable?

From the perspective of research design, my initial reaction was that it could be reasonable. But there is no denying that four unadjusted tests fail to control the family-wise Type I error rate. This is a gray area in practice: simple-effects tests are often reported without adjustment in empirical research. A more defensible—and usually more persuasive—approach is to control the error rate with a procedure such as Bonferroni, thereby treating the set of comparisons as a family of post-hoc tests. So although, in principle, a research hypothesis should be tested with a planned comparison, in practice we may need a post-hoc correction when we conduct too many tests.

But perhaps that is not the main issue here. We have described a "post-hoc" set of four comparisons, adjusted with Bonferroni: $A-B$, $A-C$, $A-D$, and $A-E$. If these were genuinely the post-hoc comparisons following some omnibus test, what would that original test be?

The usual post-hoc procedure for a one-way ANOVA examines all pairwise mean differences among the factor's levels. For five groups, the complete set is:

$$A-B,\; A-C,\; A-D,\; A-E,\; B-C,\; B-D,\; B-E,\; C-D,\; C-E,\; D-E.$$

In other words, it enumerates every pair of levels and tests whether their means differ.

Our four comparisons—$A-B$, $A-C$, $A-D$, and $A-E$—do not enumerate every pairwise difference. Is there an omnibus hypothesis whose complete set of pairwise comparisons consists of exactly these four tests?

Look just above: we already encountered a candidate. For the five groups $A$ through $E$, consider the contrast $(4,-1,-1,-1,-1)$. Could the four differences between $A$ and each treatment group be viewed as the complete set of pairwise comparisons associated with that planned contrast?

Here, then, is the question I genuinely want to leave open: in the study described at the beginning, may the researcher focus only on the four control-versus-treatment comparisons? Or should they first test the contrast $(4,-1,-1,-1,-1)$ and then proceed to the pairwise comparisons? What do you think?
:::


Let's test our contrast matrix in R:

```{r}
df4.1$C1 <- ifelse(df4.1$iden == "Plastic Bag", 1, 
                   ifelse(df4.1$iden == "Blank Person", -2, 1))
df4.1$C2 <- ifelse(df4.1$iden == "Attack Helicopter", 1, 
                   ifelse(df4.1$iden == "Blank Person", 0, -1))

my.lm4.6 <- lm(magic_score ~ C1 + C2, df4.1)
summary(my.lm4.6)
```

Examining the regression coefficients:

```{r}
group_means
```

```{r}
compr_value1 <- group_means[group_means$iden == "Plastic Bag", "magic_score"] +
                group_means[group_means$iden == "Attack Helicopter", "magic_score"] -
                2 * group_means[group_means$iden == "Blank Person", "magic_score"]
compr_value1
```

```{r}
compr_value1 / 6
```

The coefficient $b_1 = -2.034$ equals the contrast value divided by 6 ($\sum w_i^2 = 1^2 + 1^2 + (-2)^2 = 6$). Dividing by a constant scalar does not alter the $t$-statistic or $p$-value.

Now, consider what happens if we chose a different second contrast: comparing Bag People directly to Blank People:

| Identity | $C_1$ | $C_{2n}$ |
|---|:---:|:---:|
| Helicopter | 1 | 0 |
| Bag | 1 | 1 |
| Blank | -2 | -1 |

```{r}
df4.1$C2n <- ifelse(df4.1$iden == "Attack Helicopter", 0, 
                    ifelse(df4.1$iden == "Blank Person", -1, 1))

my.lm4.7 <- lm(magic_score ~ C1 + C2n, df4.1)
summary(my.lm4.7)
```

Look at the coefficient for $C_1$: it jumped from $-2.034$ to $-9.062$, even though we did not alter the coding for $C_1$ at all! Why did $C_1$'s value morph?

#### Orthogonality of Contrast Variables

This introduces the golden rule of contrast coding: contrast vectors must be **orthogonal**.

Two contrast vectors $C_1$ and $C_2$ are orthogonal if and only if:
1. The weights within each contrast sum to zero: $\sum w_{1j} = 0$ and $\sum w_{2j} = 0$.
2. The sum of their pairwise cross-products equals zero:[^chap4-11]

$$\sum_{j=1}^{k} w_{1j} w_{2j} = 0$$

Check our first contrast coding:
- Helicopter: $1 \times 1 = 1$
- Bag: $1 \times (-1) = -1$
- Blank: $(-2) \times 0 = 0$
- Sum: $1 + (-1) + 0 = 0$. **Orthogonal!**

Now check our flawed second contrast coding:
- Helicopter: $1 \times 0 = 0$
- Bag: $1 \times 1 = 1$
- Blank: $(-2) \times (-1) = 2$
- Sum: $0 + 1 + 2 = 3 \ne 0$. **Non-orthogonal!**

When contrasts are non-orthogonal, collinearity entangles their regression slopes: the estimate for $b_1$ becomes contaminated by $b_2$. Orthogonality guarantees that each contrast partitions an independent, additive slice of sum of squares without mutual interference.[^chap4-12]

The algebra makes this contamination visible. For the successful orthogonal coding, the group predictions satisfy:

$$\begin{align}
\hat{y}_{\text{Bag}} &= b_0 + b_1 + b_2 \\
\hat{y}_{\text{Heli}} &= b_0 + b_1 - b_2 \\
\hat{y}_{\text{Blank}} &= b_0 - 2b_1 + 0b_2
\end{align}$$

Adding the first two equations and subtracting twice the third gives:

$$\hat{y}_{\text{Heli}} + \hat{y}_{\text{Bag}} - 2\hat{y}_{\text{Blank}} = 6b_1$$

The unwanted $b_2$ terms cancel because the contrasts are orthogonal. Under the non-orthogonal coding, the same calculation instead produces:

$$\hat{y}_{\text{Heli}} + \hat{y}_{\text{Bag}} - 2\hat{y}_{\text{Blank}} = 6b_1 + 3b_2$$

The coefficient we intended to test is therefore contaminated by the other contrast. In more intuitive language, the two comparisons are entangled: the comparison between Bag and Blank People partly overlaps with the comparison between the two traditional groups and Blank People. To make a contrast coefficient directly test the hypothesis it was designed to represent, the other contrast variables must not share information with it.




::: {.callout-tip icon="false"}
## Food for Thought: How Do We Find Orthogonal Codes, and How Many Are There?

I imagine that many readers' heads are spinning at the requirement that coding variables be orthogonal. It is all very well to say that the coding variables must be orthogonal, but after constructing the coding variable for the hypothesis I care about, do I also have to find another variable orthogonal to it? How many of my remaining brain cells will that take?

I have several pieces of good news. First, although planned comparisons are flexible, I have rarely seen them in empirical research (chuckles). Researchers often use post-hoc comparisons to indirectly describe the result they had in mind. For example, they may report that Helicopter People differ significantly from Blank People, Bag People differ significantly from Blank People, but Helicopter and Bag People do not differ significantly, and then conclude that Helicopter and Bag People together differ from Blank People.

It is important to note that these three post-hoc tests do not test the same hypothesis as our planned contrast $(1,1,-2)$. I do not recommend that workaround. Strictly speaking, if this is your hypothesis, you should conduct the planned comparison. Quite apart from the difference in the hypotheses themselves, this post-hoc approach also sacrifices statistical power.

The second piece of good news is that when the factor has only a few levels, there are not many possible orthogonal coding systems. As summarized on [this webpage](https://www.southampton.ac.uk/~cpd/anovas/datasets/Orthogonal%20contrasts.htm){.uri}, a three-level factor has only one orthogonal coding system—the one introduced above. A four-level factor has three, and a five-level factor has four. I cannot recall seeing an empirical study that used planned comparisons for a categorical variable with more than five levels.

Still, you never know. Suppose you had magical-talent data for every country in the world and wanted to compare average magical talent across continents.[^chap4-13] Even then, there are strategies for constructing orthogonal codes. Recall that orthogonality means independence, without overlap or entanglement. A convenient strategy is to avoid making a new contrast overlap with an earlier one. For example, after comparing Helicopter and Bag People together against Blank People, do not then compare Blank People separately with Helicopter or Bag People. Instead, construct the next coding variable to compare groups *within* the Helicopter-and-Bag subset.

Even with that convenient strategy, constructing orthogonal codes for several continents could consume a lot of brain cells. Another option is to use the currently fashionable tool—AI (see, I am not opposed to using AI). Here is a prompt I just tried:

> I have a categorical variable with 12 levels and want to use contrast coding to test the difference between the first three levels and the remaining nine levels, using the contrast $(6,6,6,-2,-2,-2,-2,-2,-2,-2,-2,-2)$. Please generate the remaining 10 coding variables so that they satisfy the orthogonality requirements and can be entered directly into a regression model for planned-contrast hypothesis tests.

The AI generated the following coding table. Its response was too long to reproduce in full:

| Group | $L_1$ (Primary contrast) | $L_2$ | $L_3$ | $L_4$ | $L_5$ | $L_6$ | $L_7$ | $L_8$ | $L_9$ | $L_{10}$ | $L_{11}$ |
|:---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|---:|
| 1 | 6 | 1 | 1 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| 2 | 6 | -1 | 1 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| 3 | 6 | 0 | -2 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | 0 |
| 4 | -2 | 0 | 0 | 1 | 1 | 0 | 0 | 0 | 0 | 1 | 1 |
| 5 | -2 | 0 | 0 | -1 | 1 | 0 | 0 | 0 | 0 | 1 | 1 |
| 6 | -2 | 0 | 0 | 0 | -2 | 0 | 0 | 0 | 0 | 1 | 1 |
| 7 | -2 | 0 | 0 | 0 | 0 | 1 | 1 | 0 | 0 | -1 | -1 |
| 8 | -2 | 0 | 0 | 0 | 0 | -1 | 1 | 0 | 0 | -1 | -1 |
| 9 | -2 | 0 | 0 | 0 | 0 | 0 | -2 | 0 | 0 | -1 | -1 |
| 10 | -2 | 0 | 0 | 0 | 0 | 0 | 0 | 1 | 1 | 0 | -1 |
| 11 | -2 | 0 | 0 | 0 | 0 | 0 | 0 | -1 | 1 | 0 | -1 |
| 12 | -2 | 0 | 0 | 0 | 0 | 0 | 0 | 0 | -2 | 0 | -1 |

Is this coding table correct? Can you work out what strategy the AI used to generate it?

As I said at the beginning of the book, in 2025, using AI responsibly presupposes that you can check whether its work is correct. AI can help simplify tedious, procedural tasks; for now, however, it cannot replace your own thinking and judgment. If you do not understand the material we have discussed, you cannot tell whether the AI-generated result meets your requirements.[^chap4-14]
:::

::: {.callout-note icon="false"}
## Extension: Contrast Coding Preset in Statistical Software

Statistics has several conventional, commonly used contrast schemes. Here, "contrast coding" is meant in the broad sense we have been discussing. The names are especially confusing: some are very similar, and some are even used interchangeably. Here are a few schemes I know; if you encounter others, please let me know so I can add them.

| Scheme name | Description | Example for a four-level factor |
|---|---|---|
| Simple contrast | Equivalent to dummy coding | $(0,1,0,0)$; $(0,0,1,0)$; $(0,0,0,1)$ |
| Simple coding[^chap4-15] | The mean comparisons are like simple contrasts, but the regression intercept, as in effects coding, is the grand mean rather than the reference group's mean | $(-1/4,3/4,-1/4,-1/4)$; $(-1/4,-1/4,3/4,-1/4)$; $(-1/4,-1/4,-1/4,3/4)$ |
| Deviation contrast | Equivalent to effects coding | $(-1,1,0,0)$; $(-1,0,1,0)$; $(-1,0,0,1)$ |
| Helmert coding[^chap4-16] (orthogonal) | Starting with the first group, each contrast compares that group with all the groups that follow | $(3,-1,-1,-1)$; $(0,2,-1,-1)$; $(0,0,1,-1)$[^chap4-17] |
| Difference contrast (also called reverse Helmert; orthogonal) | The reverse of Helmert coding | $(-1,1,0,0)$; $(-1,-1,2,0)$; $(-1,-1,-1,3)$ |
| Repeated contrast | Compares each preceding group with the next group | $(1,-1,0,0)$; $(0,1,-1,0)$; $(0,0,1,-1)$[^chap4-18] |
| Forward difference contrast[^chap4-19] | Interestingly, it may produce an effect similar to repeated contrasts, but the coding is different and is not orthogonal | $(3/4,-1/4,-1/4,-1/4)$; $(1/2,1/2,-1/2,-1/2)$; $(1/4,1/4,1/4,-3/4)$ |
| Backward difference contrast | The corresponding reverse-direction scheme | $(-3/4,1/4,1/4,1/4)$; $(-1/2,-1/2,1/2,1/2)$; $(-1/4,-1/4,-1/4,3/4)$ |
| Polynomial contrast/coding (orthogonal) | Used to test nonlinear trends in a categorical predictor; this will be discussed in a future chapter | Depends on the number of levels |

If you have carefully followed the discussion above, it is worth revisiting dummy coding and effects coding. In particular, the coefficient tests under dummy coding are planned comparisons, even though those coding columns do not satisfy the orthogonality requirement for the narrow, regression-based form of contrast coding. Many of the schemes in the table above similarly implement planned comparisons.

Contrast coding is therefore not the only way to conduct planned tests. As we saw in @tip-dcal, one can construct other coding systems to test a specific mean difference. If you are sufficiently clever, you can design a coding scheme—even a non-orthogonal one—whose coefficient tests the mean contrast you care about. The narrow form of orthogonal contrast coding is simply a convenient way for people like me, who need the algebra to be transparent, to organize the tests.

:::

A final question: what if your hypotheses are, for example, a comparison of Helicopter and Bag People together against Blank People, and a separate comparison of Bag People against Blank People? The answer is simple: construct two sets of orthogonal contrasts and test the hypotheses separately. Orthogonality is a statistical requirement; our research questions should be driven by theory. If your theoretical hypotheses happen to be orthogonal, that is a bonus.[^chap4-20]

::: {.callout-tip icon="false"}
## Food for Thought: Can Random Coding Make a Comeback for Multi-Level Categorical Variables?

In @tip-coding2, we saw that for a two-level categorical variable we do not have to use 0 and 1: any two distinct numerical values yield the same test. What happens if we try this with a multi-level categorical variable? We already know that a factor with $k$ levels needs $k-1$ coding variables to represent it in a regression model whose overall fit matches that of ANOVA.

We have also seen that coding variables can take many forms; a coding variable does not have to contain only two values. What if we assign arbitrary values to the levels of each coding variable? For our three-group example, what would happen with the following coding matrix?

| Identity | $R_1$ | $R_2$ |
|---|:---:|:---:|
| Helicopter | 1 | 233 |
| Bag | 20 | 666 |
| Blank | 300 | 4396 |

Let's try it:

```{r}
df4.1$R1 <- c(rep(1, 50), rep(20, 50), rep(300, 50))
df4.1$R2 <- c(rep(233, 50), rep(666, 50), rep(4396, 50))
my.lm2.8 <- lm(magic_score ~ R1 + R2, df4.1)
summary(my.lm2.8)
```

As expected, the overall model fit is unchanged. Interested readers can work out what hypothesis each coefficient test represents, using the same method we have used throughout this chapter.

But I have one further question. Clearly, assigning arbitrary values to all the coding variables can work, while assigning the same value to every level of a coding variable cannot. Some coding variables may repeat values across levels and still work. What precise conditions must coding variables satisfy for the regression model to reproduce the ANOVA result?
:::

The lesson is subtler than simply saying that "random coding always works." A $k$-level factor needs $k-1$ linearly independent coding columns, and the resulting design matrix must have full column rank together with the intercept. If a coding column is constant, or if the coding columns are linearly dependent, the model cannot recover all $k$ group means. When the coding columns are different numerical representations of the same $k$ categories and remain linearly independent, the model space—and therefore the omnibus fit—remains unchanged. The individual slopes, however, are transformed and should not be interpreted without deriving the corresponding linear hypotheses.

### Unbalanced Sample Sizes and Coding

In experimental psychology, balanced sample sizes ($n_1 = n_2 = \dots = n_k$) are convenient, but observational data is frequently unbalanced (e.g., female-to-male ratios in psychology majors).

Most textbook examples of multi-level categorical predictors use balanced data. A balanced sample simply means that every level of the factor contains the same number of observations. Our original dataset is balanced because each identity group contains 50 people. Real empirical studies, however, rarely achieve perfect balance. For example, teacher-training universities and psychology programs may contain many more women than men, so obtaining equal sample sizes can be expensive or impossible.

Let's artificially trim our dataset to create an unbalanced sample of 120 individuals (40 Helicopter, 50 Bag, 30 Blank):

```{r}
df4.2 <- df4.1[df4.1$ID < 131 & df4.1$ID > 10, ]
group_means2 <- aggregate(magic_score ~ iden, data = df4.2, 
                          FUN = function(x) c(mean = mean(x), n = length(x)))
group_means2
```

```{r}
head(df4.2)
tail(df4.2)
```

Does sample imbalance bias regression estimates of group means?

```{r}
my.lm4.9 <- lm(magic_score ~ D1 + D2, df4.2)
summary(my.lm4.9)
```

The estimated group means remain perfectly unbiased because OLS minimization operates on each group independently.

This does not mean that imbalance is irrelevant. The regression model can still estimate each group's mean correctly, but the meaning of the grand mean—and therefore the meaning of an effects-coded intercept or contrast—requires care.

However, a fundamental distinction emerges regarding the grand mean:

1. **Unweighted Grand Mean**: treating each group as an equal conceptual unit:

$$\bar{Y}_{\text{Unweighted}} = \frac{\bar{Y}_{\text{Bag}} + \bar{Y}_{\text{Heli}} + \bar{Y}_{\text{Blank}}}{3}$$ {#eq-unweighted}

2. **Weighted Grand Mean**: weighting each group mean by its sample size:

$$\bar{Y}_{\text{Weighted}} = \frac{n_1 \bar{Y}_1 + n_2 \bar{Y}_2 + n_3 \bar{Y}_3}{n_1 + n_2 + n_3}$$ {#eq-weighted}

```{r}
## Unweighted mean
mean(group_means2[, 2][, 1])

## Weighted mean
(group_means2[group_means2$iden == "Attack Helicopter", 2][, 1] * 40 +
 group_means2[group_means2$iden == "Plastic Bag", 2][, 1] * 50 +
 group_means2[group_means2$iden == "Blank Person", 2][, 1] * 30) / 120
```

```{r}
mean(df4.2$magic_score)
```

In standard contrast and effects coding, regression intercepts estimate the **unweighted** grand mean. In most psychological research, unweighted means are preferred because sample size discrepancies are typically sampling artifacts rather than meaningful population proportions.[^chap4-5]

#### Unweighted: The Most Common Case

If we collected 40 Helicopter People, 50 Bag People, and 30 Blank People and someone asked for the average magical talent of the people we actually observed, we would naturally compute the ordinary mean of all 120 observations. That is a weighted mean. Yet effects coding does not automatically use that quantity: its intercept is the unweighted average of the three group means. Which one is appropriate depends on the population question we are asking.

Using the weighted mean says that the unequal sample sizes themselves carry substantive information. If Bag People genuinely make up a larger share of the population, their group mean should receive greater weight when estimating the population mean. But if the unequal sample sizes merely reflect unequal sampling effort—and we believe the three underlying populations are equally large, or do not know their relative sizes—then the unweighted mean is the appropriate target. This is why unweighted comparisons are often the natural choice in psychological research.

Let's verify what the two quantities look like in our unbalanced sample:

```{r}
## Unweighted grand mean: each group contributes equally
unweighted_mean <- mean(group_means2[, 2][, 1])
unweighted_mean
```

```{r}
## Weighted grand mean: each individual contributes equally
weighted_mean <- (
  group_means2[group_means2$iden == "Attack Helicopter", 2][, 1] * 40 +
  group_means2[group_means2$iden == "Plastic Bag", 2][, 1] * 50 +
  group_means2[group_means2$iden == "Blank Person", 2][, 1] * 30
) / 120
weighted_mean
```

When the sample is balanced, the two expressions collapse to the same number. When it is unbalanced, they answer different questions. The first asks for the average of the three group means; the second asks for the average score of all observed individuals.

#### Weighted: One Attempt

When should we use a weighted mean? A classic example is stratified sampling. Suppose the sizes of age groups in our magical society differ, and we sample 1% of the population from each age group. The resulting sample sizes will differ, reflecting the groups' population proportions. In estimating the population mean, we should therefore weight each group according to its share of the population.

If that is our situation, how can we construct coding variables to test a weighted comparison? We will use the planned contrast introduced above as our starting point and derive the coding weights from the desired coefficients.

The procedure is the same logic we used throughout this chapter: decide what each coefficient should mean, write the desired group predictions in terms of those coefficients, solve the resulting linear equations, and read the coefficients back as a coding matrix. The apparently unusual fractions in the table below are not arbitrary—they are the algebraic solution to the weighted hypotheses we chose.

$$\begin{align}
b_0 &= (40\hat{y}_{\text{Heli}} + 50\hat{y}_{\text{Bag}} + 30\hat{y}_{\text{Blank}}) / 120 \\
b_1 &= (40\hat{y}_{\text{Heli}} + 50\hat{y}_{\text{Bag}}) / 90 - \hat{y}_{\text{Blank}} \\
b_2 &= \hat{y}_{\text{Heli}} - \hat{y}_{\text{Bag}}
\end{align}$$

Solving this linear system yields the weighted contrast weights:

| Identity | $CW_1$ | $CW_2$ |
|---|:---:|:---:|
| Helicopter | 1/4 | 5/9 |
| Bag | 1/4 | -4/9 |
| Blank | -3/4 | 0 |

```{r}
df4.2$CW1 <- ifelse(df4.2$iden == "Attack Helicopter", 1/4, 
                    ifelse(df4.2$iden == "Blank Person", -3/4, 1/4))
df4.2$CW2 <- ifelse(df4.2$iden == "Attack Helicopter", 5/9, 
                    ifelse(df4.2$iden == "Blank Person", 0, -4/9))

my.lm4.11 <- lm(magic_score ~ CW1 + CW2, df4.2)
summary(my.lm4.11)
```

Notice the intercept: $b_0 = 90.2292$—matching our weighted grand mean exactly!

The two slopes can be checked directly against the intended weighted comparisons:

```{r}
## Weighted contrast: Helicopter and Bag versus Blank
(40 * group_means2[group_means2$iden == "Attack Helicopter", 2][, 1] +
 50 * group_means2[group_means2$iden == "Plastic Bag", 2][, 1]) / 90 -
  group_means2[group_means2$iden == "Blank Person", 2][, 1]
```

```{r}
## Contrast between Helicopter and Bag People
group_means2[group_means2$iden == "Attack Helicopter", 2][, 1] -
  group_means2[group_means2$iden == "Plastic Bag", 2][, 1]
```

In this construction, $b_1$ is exactly the weighted contrast between the combined Helicopter-and-Bag mean and the Blank mean, and $b_2$ is exactly the Helicopter-minus-Bag difference. There is no additional scale factor in these two coefficient interpretations. The chosen coding columns are also orthogonal with respect to the observed group counts:

$$40(1/4)+50(1/4)+30(-3/4)=0$$

$$40(5/9)+50(-4/9)+30(0)=0$$

$$40(1/4)(5/9)+50(1/4)(-4/9)+30(-3/4)(0)=0$$

Thus, each column's count-weighted sum is zero, and their count-weighted cross-product is zero. The two slope tests therefore correspond directly to the two specified comparisons.

This is the practical recipe for designing weighted coding schemes: use weights that represent the target population proportions, write down what each regression coefficient should mean, solve the resulting system of equations, and then construct the coding columns. Do not use sample counts as population weights unless the sampling design justifies that interpretation.

By mastering dummy coding, effects coding, orthogonal contrast coding, and weighted systems, you can translate any substantive theoretical hypothesis directly into the General Linear Model.




























































[^chap4-1]: I have always felt that the Chinese terms for "planned" and "post-hoc" carry a slightly melodramatic, relationship-drama vibe. These are simply the established translations, though.

[^chap4-2]: You might ask why the many significance tests in a correlation matrix are often reported without a Type I error correction. It is a good question—but perhaps do not ask me next time. I can only point to the gap between ideal practice and reality.

[^chap4-3]: A reader might ask why we use two coding variables for a three-level factor rather than three. The three group means are represented by the regression coefficients, including the intercept: $\hat y_{\text{Heli}}=b_0$, $\hat y_{\text{Bag}}=b_0+b_1$, and $\hat y_{\text{Blank}}=b_0+b_2$. These are three equations for three unknown coefficients. Adding another coding variable would create four unknown coefficients for only three group means. The intercept already uses one piece of information, leaving $k-1$ coding variables for a $k$-level factor—a point closely related to degrees of freedom.

[^chap4-4]: For this example, let us suppose that Blank People are a distinct population group whose gender identity is "blank."

[^chap4-5]: Typical examples include biological sex, ethnicity, nationality, and other demographic characteristics.

[^chap4-6]: The Blank People have disrupted our original plan, so for this comparison we leave them out of the contrast itself and let them contribute only to the grand mean.

[^chap4-7]: I have joked about the algebra, but I strongly encourage readers to derive every coding scheme in this chapter by hand. Understand how group means map onto regression coefficients, and verify that each coefficient tests the hypothesis you intend. For example, substituting the first equation into the next two gives $b_1=\hat y_{\text{Bag}}-b_0$ and $b_2=\hat y_{\text{Heli}}-b_0$. Substituting these expressions into the first equation and solving yields the group-mean relationships shown in the text.

[^chap4-8]: This fictional cultural history is entirely made up. Please do not take it as a claim about real-world identities.

[^chap4-9]: Here we are discussing population means, not sample means, so sample-size weighting is not part of this theoretical comparison.

[^chap4-10]: Thanks to Dr. Yujun Li for sharing the social-media example that prompted this discussion.

[^chap4-11]: I do not know how common this misconception is. Some explanations suggest that, regardless of how many coding variables there are, it is enough for the row-wise products to sum to zero. For the regression-based contrast procedure discussed here, that is not generally sufficient. Repeated contrasts provide a useful counterexample; we will encounter their coding form later.

[^chap4-12]: A small course-in-ideology moment in a statistics book: dear readers, be honest and straightforward in your dealings; entangled relationships can harm both yourself and others.

[^chap4-13]: In a real study involving countries nested within continents, I suspect multilevel modelling would be more appropriate. As I have said, I may introduce it as a side branch if I have the time and energy.

[^chap4-14]: One more update: after asking the AI fool a few follow-up questions, I eventually obtained the coding table I wanted.

[^chap4-15]: I am not even sure whether simple coding is a distinct system from simple contrast or the same system under another name. Some authors use the term for a different coding scheme.

[^chap4-16]: I do not know a satisfying translation for Helmert coding. It is generally retained as a proper name, so please bear with it.

[^chap4-17]: The second row of the Helmert coding may also be divided by 4.

[^chap4-18]: In the table, repeated contrast is shown illustratively; check its properties before entering a particular coding directly into a regression model.

[^chap4-19]: Yes, Forward Difference Contrast has a name very similar to the preceding scheme, with a direction added, but its coding is quite different. I gave up trying to translate the name further.

[^chap4-20]: But there is no perfect study in this world. —That is my observation.

```{r}
#| echo: false
save.image("data/chapter4_env.RData")
```
